Explorative data structure comparison with application to \ldots

### 2017-03-06 article/article_v0.01.tex

In this paper we propose three diagnostic plots for comparing the loading patterns in two datasets $x_{11},\dotsc,x_{1 n_1} \in \RR^d$ and $x_{21},\dotsc,x_{2 n_2} \in \RR^d$ in the same $d$ variables. The construction of these plots, which we call \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot}, proceeds in the following steps:
\begin{enumerate}
\item Standardized the observations $x_{i1},\dotsc,x_{i n_i}$ to have zero mean and unit standard deviation separately for each of the $d$ variables within each datasets. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized observations, and form the principal component analysis
\begin{equation*}
S_i = \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$.} 
\end{equation*}
\item Combine the standardized datasets into a joint dataset $\tilde{x}_1,\dotsc,\tilde{x}_n \in \RR^d$ with $n=n_1+n_2$ observations, and form the principal component analysis
\begin{equation*}
S = \frac{1}{n-2} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top.
\end{equation*}
\item Make the diagnostic plot of interest. Details of the plots will be described below.
\item To investigate the sample variation randomly reallocate the combined dataset formed in step 2 to two datasets of $n_1$ and $n_2$ observations, respectively, and redo step 1 and 3. This is done several times, e.g.\ 10000 times, and compared to the plot for the original allocation.
\end{enumerate}

@@

In step 3 we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics given by
\begin{align*}
\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} \bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)^2.
\end{align*}
From the random reallocations in step 4 we hence get p-values for the Kolmogorov-Smirnov and the Cram\'er-von Mises Goodness-of-Fit tests. Step 4 is also used to visualize the sampling variation in a display, where we plot the observed curve together with 20 of the resampled curves as well as a shaded region visualizing pointwise 95pct coverage intervals. If the observed curve is very different from the resampled curves or if it is substantially outside the shaded region, then this also indicate differences between the two datasets.

@@

Description of the \emph{HairPlot:} Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets. The empirical correlation matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
\begin{equation*}
S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
\end{equation*}
and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ anticlockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn anticlockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k} \eta_{2j}^\top \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j} \eta_{1k}^\top \rvert, &
\theta_{jk} &= \acos(\lvert \eta_{1k} \eta_{2j}^\top \rvert).
\end{align*}

@@

In the \emph{HairPlot} the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal components for the first dataset in the coordinate system of the second dataset. Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components for the second dataset in the coordinate system of the first dataset. If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length. Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal. And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells as well as arrows of non vanishing length in the off-diagonal cells.

%Using the same grid display, but interchanging the first and the second dataset as well as the rows and the columns, and using the angles clockwise, similar arrows are plotted in a different color.

%In our implementation we have used blue for the first and the second set of arrows, respectively.

@@

Remarks: Presently permutation tests based on the Kolmogorov-Smirnov and the Cram\'er-von Mises test statistics has been implemented for CumVarAgreement. This should also be possible for the Hairplot based on the off-diagonal components. For the Pancakeplot a Wallyplot has been implemented (however, in Claus' Wally plot the position of the ``observed dataset'' is also random).

Description of the \emph{PancakePlot}: Our proposal of a PCADSC method consist of three steps. These steps should be performed separately for each of the two (or more) datasets that we wish to compare. Note that the datasets must have the same variables, but different sample sizes are allowed. The three steps are:
\begin{enumerate}
\item \hl{Standardize. Full data or subsets? And also: Something about what to do if not all variables are numerical.}
\item Compute the PCA loadings and the variance contributions of each principal component.
\item For each principal component, standardize the loadings, i.e. scale them such that they sum to one.
\item Produce a plot consisting of a bar for each principal component, decorated with the cumulative variance contribution corresponding to this component. The bar should be of length one and colored according to the variables loading the component.
\end{enumerate}
The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.  \hl{Refer to example/show plot.}

>>>>

In this paper we propose three diagnostic plots for comparing the loading patterns in two datasets $x_{11},\dotsc,x_{1 n_1} \in \RR^d$ and $x_{21},\dotsc,x_{2 n_2} \in \RR^d$ in the same $d$ variables. The construction of these plots, which we call \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot}, proceeds in the following steps:
\begin{enumerate}
\item Standardized the observations $x_{i1},\dotsc,x_{i n_i}$ to have zero mean and unit standard deviation separately for each of the $d$ variables within each datasets. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized observations, and form the principal component analysis
\begin{equation*}
S_i = \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$.} 
\end{equation*}
\item Combine the standardized datasets into a joint dataset $\tilde{x}_1,\dotsc,\tilde{x}_n \in \RR^d$ with $n=n_1+n_2$ observations, standardize this dataset, and form the principal component analysis
\begin{equation*}
S = \frac{1}{n-2} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top.
\end{equation*}
\item Make the diagnostic plot of interest. Details of the plots will be described below.
\item To investigate the sample variation randomly reallocate the combined and standardized dataset formed in step 2 to two datasets of $n_1$ and $n_2$ observations, respectively, and redo step 1 and 3. This is done several times, e.g.\ 10000 times, and compared to the plot for the original allocation.
\end{enumerate}

@@

In step 3 we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics given by
\begin{align*}
\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} \bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)^2.
\end{align*}
From the random reallocations in step 4 we get p-values for the Kolmogorov-Smirnov and the Cram\'er-von Mises Goodness-of-Fit tests. Step 4 is also used to visualize the sampling variation in a display, where we plot the observed curve together with 20 of the resampled curves as well as a shaded region visualizing pointwise 95pct coverage intervals. If the observed curve is very different from the resampled curves or if it is substantially outside the shaded region, then this also indicate differences between the two datasets.

@@

Description of the \emph{HairPlot:} This plot simultaneously compares the variation components and the loadings. Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets. The empirical correlation matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
\begin{equation*}
S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
\end{equation*}
and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ anticlockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn anticlockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k} \eta_{2j}^\top \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j} \eta_{1k}^\top \rvert, &
\theta_{jk} &= \acos(\lvert \eta_{1k} \eta_{2j}^\top \rvert).
\end{align*}

@@

In the \emph{HairPlot} the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal components for the first dataset in the coordinate system of the second dataset. Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components for the second dataset in the coordinate system of the first dataset. If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length. Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal. And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells as well as arrows of non vanishing length in the off-diagonal cells. The name \emph{HairPlot} arise by seeing the arrows as \emph{hairs}. Using this interpretation we may say that a more ``hairy'' plot indicates less agreement between the two loading patterns.

Along with the visual comparison we have also implemented numerical tests in step 3. Separately for the off-diagonal grid cells we may use the lengths $\mu_{jk}$ and $\nu_{jk}$ as test statistics, and for the on-diagonal grid cells we use the distance between the scaled loadings as test statistics, that is
\begin{equation*}
\chi_k = \lambda_{\max}^{-1} \lvert \lambda_{1k} \eta_{1k} - \lambda_{2k} \eta_{2k} \rvert, \quad \text{for $k=1,\dotsc,d$.}
\end{equation*}
In total this gives $2d^2 - d$ tests, which should be corrected for the family wise type I error using a methods that doesn't assume independence among the tests. We also construct overall Kolmogorov-Smirnov and Cram\'er-von Mises type test statistics by
\begin{align*}
\text{KS}_1 &= \max_{j,k=1,\dotsc,d} (1_{j \neq k} \mu_{jk} + 1_{j=k} \chi_k), &
\text{CvM}_1 &= \sum_{j=1}^d \sum_{k=1}^d (1_{j \neq k} \mu_{jk} + 1_{j=k} \chi_k)^2, \\
\text{KS}_2 &= \max_{j,k=1,\dotsc,d} (1_{j \neq k} \nu_{jk} + 1_{j=k} \chi_k), &
\text{CvM}_2 &= \sum_{j=1}^d \sum_{k=1}^d (1_{j \neq k} \nu_{jk} + 1_{j=k} \chi_k)^2.
\end{align*}

@@

Description of the \emph{PancakePlot}: This plot compares the loadings.

Our proposal of a PCADSC method consist of three steps. These steps should be performed separately for each of the two (or more) datasets that we wish to compare. Note that the datasets must have the same variables, but different sample sizes are allowed. The three steps are:
\begin{enumerate}
\item \hl{Standardize. Full data or subsets? And also: Something about what to do if not all variables are numerical.}
\item Compute the PCA loadings and the variance contributions of each principal component.
\item For each principal component, standardize the loadings, i.e. scale them such that they sum to one.
\item Produce a plot consisting of a bar for each principal component, decorated with the cumulative variance contribution corresponding to this component. The bar should be of length one and colored according to the variables loading the component.
\end{enumerate}
The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.  \hl{Refer to example/show plot.}
